ISYE 6740 HW 1

### 2024-08-27 msiddiqui60_HW1_report.tex
update with section 2 added

\textbf{Relevant Eigenvectors (correspond to eigenvalues = 0)}
\[
\begin{matrix}
-0.57735027 & 0  \\
-0.57735027 & 0  \\
-0.57735027 & 0  \\
0 & 0.70710678  \\
0 & 0.70710678
\end{matrix}
\]

\end{enumerate}

>>>>

\textbf{Relevant Eigenvectors (correspond to eigenvalues = 0)}
\[
\begin{matrix}
-0.57735027 & 0  \\
-0.57735027 & 0  \\
-0.57735027 & 0  \\
0 & 0.70710678  \\
0 & 0.70710678
\end{matrix}
\]
\end{enumerate}

\section{Math of k-means Clustering}

\item :et's look at the equations provided before we start to answer the questions.
\begin{equation}
J =\sum_{i=1}^m\sum_{j=1}^k r^{ij} \|\text x^i-\mu^j\|^2,
\label{J_def}
\end{equation}
$$\mu^j=\frac{\sum_i r^{ij} \text x^i}{\sum_i r^{ij}}.$$

The first equation states that the k-means function seeks to minimize the distortion J by summing up
the squared euclidian distance for each data point from its respective centroid.The r_ij term ensures that
we do not take the distance of a datapoint from a centroid other than its own in our calculation. The final
derived equation below states that the centroid Mu for a given cluster j that minimizes the distortion
function J is the sum of all data points in that cluster divided by the total number of data points in that
cluster, which is essentially the average of all the points in that cluster j. Similarly to a previous concept
question, we can first expand the squared Euclidian distance to write in a format we are accustomed to. To
find critical points such as minimum or maximum, we will then take the derivative with respect to our variable
of interest (in this case the centroid Mu_j which we are trying to optimize to minimize the distortion function
J). Once we set the derivative to 0, we can solve for the value of centroid Mu_j.

\[
\| x_i - \mu_j \|^2 = (x_i - \mu_j)^T(x_i - \mu_j)
\]

\[
\frac{\partial J}{\partial \mu_j} = \sum_{i=1}^{m} r_{ij} \cdot 2(\mu_j - x_i) = 0
\]

\[
\mu_j = \frac{\sum_{i=1}^{m} r_{ij} x_i}{\sum_{i=1}^{m} r_{ij}}
\]

In the final steps, Mu_j can be pulled out of the sum and treated as a constant as it doesn't vary with i
like the r_ij term does, allowing us to simply expand the summation terms and divide to isolate Mu_j on the
left side of the equation.

\item The variable r_ij is 1 when the datapoint x_i belongs to the jth cluster and 0 when it doesn't as
mentioned in the first part of the section. If we have a fixed centroid and need to determine whether
r_ij should be 1 or 0, we need to find the closest centroid to that data point and only assign 1 when
that is true. To accomplish this, for a given point x_i, we will need to traverse all centroids Mu_j for all
j clusters and then find the minimum distance (in our case the squared euclidian distance) between the point
and the centroid and return 1 for that combination. As in the previous concept section and lectures, we can use the
concept of argmin function to help accomplish this. Mathematically, our function will appear as:

\begin{equation}
\[
r_{ij} =
\begin{cases}
1 & \text{if } j = \arg \min_{j'} \| x_i - \mu_{j'} \|^2 \\
0 & \text{otherwise}
\end{cases}
\]
\end{equation}

This will minimize the distortion function j as the r_ij values are now set appropriately.

\item We'll now accomplish part 1 and 2 for the situation where instead of euclidean distance,
we use the Mahalnobis distance in minimizing the cost-function.
\[
J =\sum_{i=1}^m\sum_{j=1}^k r^{ij} (\text x^i-\mu^j)^T \Sigma  (\text x^i-\mu^j),
\]

Per the source I've linked below, the Mahalnobis distance is the relative distance of point x_i, from the center
of mass y (in our case, the centroid Mu_j). The transpose vector of x_i - Mu_j is multiplied by the
inverse of the covariance matrix multipled by x_i - Mu_j. This essentially introduces a normalization
component. Additionally, like our squared euclidian distance in the prior problem, we have the squared
Mahalanobis distance. The steps we take will be identical to those we took to solve for the
respective Mu_j and r_ij of the euclidian squared distance. However, now we simply have an additional
covariance term to be concerned with. I'll skip ahead to the parts that will look different.
\[
\frac{\partial J}{\partial \boldsymbol{\mu}_j} = -2 \sum_{i=1}^{m} r_{ij} \Sigma (\mathbf{x}_i - \boldsymbol{\mu}_j)
\]

\[
-2 \sum_{i=1}^{m} r_{ij} \Sigma (\mathbf{x}_i - \boldsymbol{\mu}_j) = 0
\]

\[
\sum_{i=1}^{m} r_{ij} \Sigma \mathbf{x}_i = \sum_{i=1}^{m} r_{ij} \Sigma \boldsymbol{\mu}_j
\]

\[
\boldsymbol{\mu}_j \sum_{i=1}^{m} r_{ij} = \sum_{i=1}^{m} r_{ij} \mathbf{x}_i
\]

\[
\boldsymbol{\mu}_j = \frac{\sum_{i=1}^{m} r_{ij} \mathbf{x}_i}{\sum_{i=1}^{m} r_{ij}}
\]

Note, that we 'cancelled' out the covariance definite matrix present on both sides. Also note how the
optimal centroid M_j for cluster j looks identical to the one from the l2-norm squared, as the average
of the data points within the cluster. For the cluster assignment solving for r_ij, we'll follow the same
logic we did for the l2_norm squared distance. Namely, we realize we are setting r_ij to 1 where
the distance between a point and cluster is the minimum for all given clusters. We will simply replace
the distance function to the Mahalanobis distance squared formula provided in the assignment.

\[
r_{ij} = \begin{cases}
1 & \text{if } j = \arg \min_{j'} \; (\mathbf{x}_i - \boldsymbol{\mu}_{j'})^T \Sigma (\mathbf{x}_i - \boldsymbol{\mu}_{j'}) \\
0 & \text{otherwise}
\end{cases}
\]

Source used to understand Mahalanobis distance: https://math.stackexchange.com/questions/428064/distance-of-a-test-point-from-the-center-of-an-ellipsoid
Another source: https://www.statisticshowto.com/mahalanobis-distance/
Another source: https://mccormickml.com/2014/07/22/mahalanobis-distance/
\end{enumerate}

\section{Image Compression Using Clustering}
